---
title: "pH in the Poudre"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
editor_options: 
  chunk_output_type: console
---

# Background


This week's content covers acid base chemistry, so to tie that 
work to real data we will be working with pH and 
chemistry data from the Poudre River. 

```{r setup, include = F}

# You don't have to understand any of this, it's just here 
# for those who are interested!

library(dataRetrieval) #Package that retrieves public water quality data
library(tidyverse) #Package for making graphs and manipulating data
library(dygraphs) #Make fancy plots
library(xts) #Time series graph

## Setup pretty options
knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment = FALSE)
```

## Why work in R? 

This assignment shows an example where working with R can really 
make the work go faster, be more beautiful, and have a bigger
impact on our audience. 

### Downloading data

Here we will use the `dataRetrieval` package to download some pH
data from the United States Geological Survey who took daily pH
measurements in the Poudre for all of the 1990s. The USGS
station is right by Odell Brewing Co for reference.

If you haven't worked with usgs data before then what you need to is 
that each USGS station (where they measure discharge and other 
water parameters like pH) has a unique siteID code. Each possible
measurement also has a unique parameter code. So we will need 
this info here

Site ID = `06752260`
Parameter Code for pH = `00400`
Parameter Code for Q (discharge) = `00060`
startDate = `1989-01-01`
endDate = `1999-12-31`

We will use the function `readNWISdv` to download data with these
parameters

```{r, cache = T}

# Download pH data and rename the columns
ph <- readNWISdata(siteNumbers = '06752260',
                 startDate = '1989-01-01',
                 endDate = '1999-12-31',
                 parameterCd = '00400') %>%
  select(dateTime,ph = X_00400_00002)


#Download Discharge data for same period
q <- readNWISdata(siteNumbers = '06752260',
                 startDate = '1989-01-01',
                 endDate = '1999-12-31',
                 parameterCd = '00060') %>%
  select(dateTime,q_cfs = X_00060_00003)

```


### Exploring the data

Now that we have downloaded both river discharge and pH time
series we may want to take a look at these data. 

A few useful functions for looking at data are:
`View`, `head`, and `str`

#### View

```{r}
## Your code Here

# View(ph)


```

#### head

```{r}
head(ph)

```

#### str (Structure)

Ctrl + Alt + I

```{r}
str(q)

```


### Plotting the data



#### Static Plot

```{r}

ggplot(ph, aes(x = dateTime, y = ph)) + 
  geom_point() +
  xlab('Date') + 
  ylab('pH')

```

#### Dynamic plot

```{r}
ggplot(q, aes(x = dateTime, y = q_cfs)) + 
  geom_line() + 
  scale_y_log10() + 
  xlab('Date') + 
  ylab('Q (cfs)')

q_ts <- xts(q$q_cfs, order.by = q$dateTime)

dygraph(q_ts)
```


# Assignment

Now we will use the data we just downloaded and manipulated to explore how pH varies in the Poudre and what might control
pH. 



## Q1 - Controls on pH

Make a plot of Q versus pH

#### Joining the data

```{r}

# Joining Q and pH data to make a single data frame
qph <- left_join(q, ph, by = 'dateTime')

qph_clean <- filter(qph, ph > 2.5)
```

#### Plotting Q vs pH


```{r}

```


## Q2 What might be controlling pH in the Poudre? 

TYPE YOUR ANSWER HERE


## Q3 Make a bonus plot

```{r}

```

