---
title: "Pretty Report"
author: "Nick Gubbins"
date: "4/4/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dataRetrieval)
library(mapview)
library(lfstat)
library(sf)
library(here)
```

# Sharing Data

A key part of the scientific process is communicating your findings with others. 
While many of you have had experience doing this in more formal settings 
(like an essay, poster, or talk), often times scientists need to communicate 
intermediate results to collaborators and stakeholders. Knitted R markdown documents are a great option for this kind of communication. HTML documents allow for high quality, interactive figures with relatively little initial investment. 

# Assignment

In this assignment you will go back to our four previous assignments and 
recreate three figures as beautifully as possible. Your final product will be a report that is clear of errors and easy to interpret. Imagine this report is going to be shared with a group of students interested in what we are doing in this course.

For each plot you will:

* Recreate a previous figure

    * Copy and paste the original code from your past assignment

    * Move any underlying data from the assignment to this project's data folder
    
    * Load any packages needed for the figure

    * Recreate/read in any data objects needed to make the figure
    
* Write a short paragraph on how the figure will be improved

    * Cite specific design principles from our guest lecturer
    
    * Explain why previous design choices were ineffective
    
    * Explain how your new design choices will rectify those issues
    
* Implement your written changes into a new, improved figure

Each plot should:

* Make it clear where the data was collected and who collected it

* Have a design that clearly highlights the effect of interest

* Include a descriptive caption

* Have proper labels and units

* Have a helpful title (not just 'Y vs X')

**The goal of this assignment is to have the cleanest, prettiest, most effective knitted document you can make.** 

**Delete all my introductory text, check for errors, and make sure your name is listed as the author before submitting.**

**Take pride in it and make it to be shared with others!**

## Figure 1

### Figure as it existed (copy paste code here):

```{r}

```

### Describe how you will improve this figure:

### Improved figure

```{r}

```

## Figure 2

### Figure as it existed (copy paste code here):

```{r}

```

### Describe how you will improve this figure:

### Improved figure

```{r}

```


## Figure 3

### Figure as it existed (copy paste code here):

```{r}

```

### Describe how you will improve this figure:

### Improved figure

```{r}

```

