---
title: "Dissolved Oxygen in the Cache La Poudre"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse) # Data wrangling and viz
library(mapview) # For interactive maps
library(sf) # Spatial data
library(lubridate) # Dealing with time
library(dygraphs) # dynamic graphs
library(xts) # time series
#remotes::install_github("r-spatial/mapview")



knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, 
                      comment = F)
```

# Background

In late summer 2018, there was a large fish kill in the Cache La Poudre river in Fort Collins, Colorado. After that event, the City of Fort Collins and the 
Ross lab installed 6 water quality sensors in the river to get real-time water 
quality information. One key parameter these sensors measured was dissolved 
oxygen, because fish kills can be caused by low oxygen levels, sometimes called
anoxic events. These events are caused by a variety of things, but often it is when water is slow moving, has a highly productive algal community in the river (or lake). These algae produce oxygen during the day but at night and when the 
waters cool, they die. Their decomposing bodies consume oxygen rapidly. 

For this assignment we will explore controls on dissolved oxygen at a variety
of sites in the Cache La Poudre. 

## Site Map

```{r}
sites <- st_read('data/sondes.kml') %>%
  filter(Name != 'Manners Sonde (FC-existing)') %>%
  mutate(Name = factor(Name,levels=c('Rist Canyon Rd (CSU)',
                                     'Legacy (CSU)',
                                     'Lincoln (In-situ)',
                                     'Timberline (FC)',
                                     'Prospect (FC)',
                                     'ELC (CSU; online)',
                                     'Archery Range (FC)',
                                     'Fossil Creek (FC)')))


mapview(sites, zcol = 'Name', col.regions=c('#1b7837','#7fbf7b','gray',
                                        '#d9f0d3','#e7d4e8',
                                        '#af8dc3','#762a83','gray'
                                        ))
```

### Question 1)

When thinking about controls on dissolved oxygen in rivers, what are three 
important gradients that change as you go from Rist Canyon to Archery Range? 
(hint use the different basemaps to explore the landscape)

TYPE YOUR ANSWER HERE (3-5 Sentences)


## Dataset 

For this project the City and the Ross Lab are measuring:

- Temperature
- Dissolved Oxygen (as % saturation and mg/l)
- Turbidity
- pH
- ORP (oxidation reduction potential)
- Water Level
- Specific Conductivity (salinity)

Here we will be looking at a subset of this sensor data for October, 2019.

```{r}

p_dat <- read_csv('data/oct_2019.csv')


```


## Site specific analysis


### Rist Canyon


```{r}
rist_dat <- p_dat %>%
  filter(site == 'Rist',
         day(hour) > 7)

ggplot(data = rist_dat, aes(x = hour, y = value)) + 
  geom_line() + 
  facet_wrap(~sensor, scales = 'free')
```


### Reshaping data

```{r}
cool_sensors <- c('depth','do_mg','salinity','temp','turb')

rist_clean <- rist_dat %>%
  filter(sensor %in% cool_sensors)

ggplot(data = rist_clean, aes(x = hour, y = value)) + 
  geom_line() + 
  facet_wrap(~sensor, scales = 'free',ncol = 1)

rist_wide <- pivot_wider(rist_clean,
                         names_from = sensor,
                         values_from = value)

rist_xts <- xts(rist_wide %>% select(do_mg,depth),
                order.by = rist_wide$hour)

dygraph(rist_xts)
```


## Question 2) Pick a different site and make a graph with 4 sensor timeseries (must include DO)

```{r}
# Your Code here
```


## Question 3) With the sensors you chose, what correlations do you see between different datasets? 

## Question 4) Make a plot with your 4 sensor time series and data for Rist Canyon and your site.


```{r}
prospect_rist <- p_dat %>%
  filter(site == c('Prospect', 'Rist'),
         day(hour) > 7)
```


## Question 5) What landscape variable differences might explain
the differences you see in your plot? 


## Question 6) What River morphology (shape) differences might
explain differences you see in your plots? 


## Question 7) Make a plot for all sites with DO on the X-axis
and pH on the Y

```{r}

## Your code here
```


## Using your google skills, describe the relationship between
pH and dissolved oxygen, why/how are these variables related? 


