---
title: "Temperature and DO"
author: "Nick Gubbins"
date: "1/26/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dataRetrieval)
library(lubridate)
library(tidyverse)
```

# Ecological Context

## Temperature as a master control

Available energy in a system is the master control on which, and at what rate, 
chemical reactions take place. Temperature is the most common, and most easily 
accessible, measurement of energy in a system. While other forms of energy (
such as turbulence) are critical for understanding aquatic systems, temperature 
can be easily and cheaply measured without effecting the system itself. 
For this week's assignment, we will be getting more familiar with R by 
downloading and analyzing temperature data from the USGS.

## Power generation and water temperature

Rivers are often used as a water source for the production of steam in power 
generation. Generating stations draw in water from upstream, use it to create 
steam, use that steam to spin turbines, then (mostly) cool the steam, and 
re-release it downstream. You can clearly see the effect of this on the 
White River in downtown Indianapolis, IN. Station 03353000 is located 
just upstream of a major generating plant. Station 03353611 is just below it.
The gauges are only ~2 miles apart. Let's load in a few months of data from
both and see how they differ.

```{r}
ind <- data <- readNWISuv(siteNumbers = c('03353611', '03353000'),
                   parameterCd = '00010',
                   startDate = '2021-10-01',
                   endDate =  '2022-01-01')

ggplot(ind, aes(x = dateTime, y = X_00010_00000, color = site_no))+
  geom_line()+
  labs(x = 'Date', y = 'Temperature (degrees C)', color = 'Site')
```

You can see the effect of the generating station most clearly as temperatures
drop for the winter. Water is usually returned at ~20 degrees C, giving the 
downstream site a higher average temperature. 

## Dissolved oxygen and temperature

Dissolved oxygen is a key indicator of aquatic health. It is produced by 
photosynthesizing microbes and macrophytes. Heterotrophs, like fish and many 
kinds of bacteria, use dissolved oxygen in respiration. Healthy freshwater 
systems tend to have dissolved oxygen levels above 6 mg/L (usually around 80% 
saturation or higher). When dissolved oxygen levels get too low (below 6 mg/L) 
fish cannot spawn, as fish eggs rely on passive diffusion of dissolved oxygen 
to survive. Below 5 mg/L DO, you can see dramatic [fish kills](https://apnews.com/article/837ec91d3a1c4f86b1258be93cd865e8). These kills represent a huge disruption to the river's food chain and a massive 
influx of organic matter, often leading to spiraling water quality issues.

# Assignment

## Q1:
Read in **mean daily** discharge (code '00060'), DO (code '00300'), 
and temperature (code '00010') data from the South Platte River at Englewood, CO
('06711565') for the 2019 water year.

```{r}
data <- readNWISdv(siteNumbers = '06711565',
                   parameterCd = c('00060', '00300', '00010'),
                   startDate = '2018-10-01',
                   endDate =  '2019-9-30')
```

## Q2: 
Look at the structure of the data in any way you choose.

```{r}
glimpse(data)
```

## Q3: 
Rename your columns to something sensible, omit rows without complete
observations, and reduce your data frame to only include the date and your data
of interest. 

```{r}
data <- data %>%
  rename(temp_C = X_00010_00003, 
         q_cfs = X_00060_00003,
         do_mgl = X_00300_00003) %>%
  select(Date, temp_C, q_cfs, do_mgl) %>%
  na.omit()
```

## Q4: 
Plot a time series of all three parameters the same graph.

```{r}
ggplot(data, aes(x = Date))+
  geom_line(aes(y = q_cfs))+
  geom_line(aes(y = temp_C))+
  geom_line(aes(y = do_mgl))
```

## Q5: 
This time series graph is not very useful to us.
Why not? Give two examples of design choices you could make to improve its
usefulness.

*The scale of discharge is way beyond that of temperature and DO. One way to improve this is to facet wrap by parameter. Another low-haning improvement would be to label the parameters by color.*

## Q6: Implement one of your proposed changes.

```{r}
data_long <- pivot_longer(data, cols = c(q_cfs, do_mgl, temp_C),
                          names_to = 'var', values_to = 'val')

ggplot(data_long, aes(x = Date, y = val, color = var))+
  geom_line()+
  facet_wrap(vars(var), ncol = 1, scales = 'free')
```

## Q7: 
Scatterplots are a great way to compare correlations. 
Create two scatterplots. One of DO vs Q, the other of DO vs temperature.

```{r}
ggplot(data, aes(x = q_cfs, y = do_mgl))+
  geom_point()

ggplot(data, aes(x = temp_C, y = do_mgl))+
  geom_point()
```

## Q8: 
Use the 'cor()' function to test which two variables are the most closely
related (look at the 'help' window for documentation). Which relationship is it?
Why do you think that is? Report your answers in complete sentences.

```{r}
cor(data$temp_C, data$do_mgl)
cor(data$q_cfs, data$do_mgl)
```

*Temperature is more closely correlated to DO than discharge. This is likely due to the lower solubility of gas in warm water and the energetics needed for photosynthesis.*

## Q9:
Imagine the city of Englewood is exploring building a coal-fired power
plant using the South Platte as its water source and you have been hired on as a
consultant. **Write a paragraph** brief for the city. Given that the benefits to
the city are concrete, what issues would you raise? What effects do you think
the plant would have on water temperatures and water quality?
What data/assurances would you need to see before thinking the project
should go ahead?

