---
title: "Ion Sorption"
output: html_document
editor_options: 
  chunk_output_type: console
---



```{r setup}

library(sf)
library(tidyverse)
library(dataRetrieval)
library(mapview)
library(lubridate)
library(ggthemes)
```


# Suspended sediment and contaminants

For this assignment we will be looking at the correlation between various
metals and suspended sediment concentrations at sites across Colorado. 




# Site Selection

```{r}

#read in spatial
sites <- st_read('sites.gpkg')


mapview(sites, legend = F)
```


## Q1) Pick a site to focus on based on random preference. Make a map of your site.


```{r}
## Using the filter function 


```



## Q2) Plot a time series of TSS data at this site, what do you think controls
variation in TSS? 


```{r}

metals_tss <- read_csv('c_metals_tss.csv')

tss <- metals_tss %>%
  filter(parameter == 'Total suspended solids' | parameter ==  'Suspended Sediment Concentration (SSC)')

## Don't use this site
tss_site <- tss %>%
  filter(SiteID == 'USGS-06754000')

# Make a plot
ggplot(tss_site %>%
         filter(year(date) > 2015), aes(x = date, y = harmonized_value)) + 
  geom_point() + 
  scale_y_log10()
```


## Q3) Plot a time series of 3 metals at this site. What do you think controls
the concentration of metals?

(if you have less than three, plot as many you can)

```{r}

```



## Q4) Make a dataset with TSS and Metals concentrations combined and plot TSS vs
Concentration. 

## Q5) What might cause these relationships? Use language from the lectures to 
justify your response




BONUS

Download Q data and look at teh correlation between Q and TSS and Q and Metals.

What might be the reason for the correlation or lack thereof? 


